Fare Calculator

### 2022-02-09 r-package/tests_marcus/fare_calculator.Rmd
Added function `dropFareCalculator()` to R5RCore Added fareCap to FareStructure Atomic values from R are not converted to JSONArrays anymore Fixed bugs in RuleBasedFareCalculator logic Updated Fare Calculator draft vignette

devtools::load_all(".")
library(sf)
library(data.table)
library(ggplot2)
library(dplyr)
library(mapview)
```

@@

To begin tackling this challenge, we are proposing a simple generic rule-based fare calculator that can be configured via a predefined set of properties and rules that can be set directly from R or using external tools such as text editors and spreadsheets. By no means we intend for this to be a robust system that can take into consideration all public transport systems out there. That would be a Herculean task, as can be seen by the long discussions taking place in the Fares V2 forums. The features included in r5r's fare calculator are inspired by our empirical observations of Brazilian public transport systems, and is meant to be used mainly in our own Access to Opportunities project [link]. Everyone else is welcome to use it, if the current features suits their needs.

@@

- Each bus ticket costs R$ 4.80.
- Riding a second bus adds R\$ 2.40 to the total cost. Subsequent bus rides cost the full ticket price of R$ 4.80.
- Each train ticket costs R$ 4.50. Once the user enters a train station, they can take an unlimited amount of trains as long as they do not leave the station. 
- The integrated tariff between bus and train has a 10% discount, which totals R$ 8.37.

@@

In the example below, the `base_fare` is the standard bus ticket price of R$ 4.80. Prices, however, must be represented as integer values, so here we are saying that the `base_fare` is equivalent to 480 cents. We are also stating that `by = "MODE"`, so that each transport mode has its own fares and integration rules. `by` can also be set to `"AGENCY_ID"` and `"AGENCY_NAME"`, so different rules can be applied to routes ran by different agencies, and to `"GENERIC"` when the entire system follows the same rules.

@@

```{r}
fare_settings$base_fare
fare_settings$max_discounted_transfers
fare_settings$transfer_time_allowance <- 60
```

@@

We need to do two small changes in the `fare_per_mode` table, specifically in the `"RAIL"` mode: setting `unlimited_transfers` to `TRUE`, and updating fare to 450. The `"BUS"` mode does not need any change, so we're leaving it with the default values. We'll do those changes below, using data.table notation.

```{r}
fare_settings$fare_per_mode[mode == "RAIL", unlimited_transfers := TRUE]
fare_settings$fare_per_mode[mode == "RAIL", fare := 450]
```

Checking the results below, everything looks ok:

```{r}
fare_settings$fare_per_mode
```
### Transfer fares configuration

@@

- The fare for "BUS" to "BUS" integration is composed of 480 for the first leg plus 240 for the second leg, which equals to a total fare of 720. Also, there is a special rule that the discount is not valid when transferring between buses within the same route (for example, from bus T1 to another T1), so we need to set `allow_same_route` to `FALSE`.

```{r}
fare_settings$fare_per_transfer[leg1 == "BUS" & leg2 == "BUS", 
                                allow_same_route := FALSE]

fare_settings$fare_per_transfer[leg1 == "BUS" & leg2 == "BUS", 
                                fare := 720]
```

- Transfers between "RAIL" and "RAIL" are free and unlimited, which is already accounted for in the field `unlimited_transfers` of the `fare_per_mode` table. Thus, the equivalent line of the `fare_per_transfer` data.frame can be completely removed.

@@

Now, we can answer questions like "how many jobs one can access in 60 minutes using public transport, on a $5 budget?".

@@

```

```{r}
r5r_core$setFareCalculatorDebugOutput(
  here::here("tests_marcus", "fare_calculator_output.csv"))
```

```{r}
access_f <- accessibility(r5r_core,
                        origins = points_small,
                        destinations = points_small,
                        departure_datetime = departure_datetime,
                        opportunities_colname = "unit",
                        mode = c("WALK", "TRANSIT"),
                        cutoffs = c(30, 45),
                        fare_calculator_settings = fare_settings,
                        max_fare = 500,
                        max_trip_duration = 45,
                        max_walk_dist = 800,
                        time_window = 1,
                        percentiles = 50,
                        verbose = FALSE)
```

```{r}
debug_output <- fread(here::here("tests_marcus", "fare_calculator_output.csv"))

debug_output[, count := stringr::str_count(debug_output$pattern, 
                                           pattern = "\\|") + 1]
```

```{r}
debug_output[stringr::str_starts(debug_output$pattern, "LINHA")] |> View()
```

```{r}
View(fare_settings$fare_per_transfer)
```

>>>>

devtools::load_all(".")
library(tidyverse)
library(sf)
library(data.table)
library(ggplot2)
library(dplyr)
library(mapview)
```

@@

To begin tackling this challenge, we are proposing a simple generic rule-based fare calculator that can be configured via a predefined set of properties and rules that can be set directly from R or using external tools such as text editors and spreadsheets. By no means we intend for this to be a robust system that can take into consideration all public transport systems out there. That would be a Herculean task, as can be seen by the long discussions taking place in the Fares V2 forums. The features included in r5r's fare calculator are inspired by our empirical observations of Brazilian public transport systems, and is meant to be used mainly in the (Access to Opportunities project)[https://www.ipea.gov.br/acessooportunidades/en/]. Everyone else is welcome to use it, if the current features suits their needs.

@@

-   Each bus ticket costs R\$ 4.80.
-   Riding a second bus adds R\$ 2.40 to the total cost. Subsequent bus rides cost the full ticket price of R\$ 4.80.
-   Each train ticket costs R\$ 4.50. Once the user enters a train station, they can take an unlimited amount of trains as long as they do not leave the station.
-   The integrated tariff between bus and train has a 10% discount, which totals R\$ 8.37.

@@

In the example below, the `base_fare` is the standard bus ticket price of R\$ 4.80. Prices, however, must be represented as integer values, so here we are saying that the `base_fare` is equivalent to 480 cents. We are also stating that `by = "MODE"`, so that each transport mode has its own fares and integration rules. `by` can also be set to `"AGENCY_ID"` and `"AGENCY_NAME"`, so different rules can be applied to routes ran by different agencies, and to `"GENERIC"` when the entire system follows the same rules.

@@

```{r}
fare_settings$base_fare
fare_settings$max_discounted_transfers
fare_settings$transfer_time_allowance <- 60
fare_settings$fare_cap
```

@@

We need to do two small changes in the `fare_per_mode` table, specifically in the `"RAIL"` mode: setting `unlimited_transfers` to `TRUE`, and updating fare to 450. We also need to change one setting in the `"BUS"` mode. There is a discount for transfer between buses (which is set in the following section), but that discount is not valid when transferring between buses within the same route (for example, from bus T1 to another T1). We can use the `allow_same_route_transfer` parameter of the `"BUS"` mode for that. We'll do those changes below, using data.table notation.

```{r}
fare_settings$fare_per_mode[mode == "RAIL", unlimited_transfers := TRUE]
fare_settings$fare_per_mode[mode == "RAIL", fare := 450]
fare_settings$fare_per_mode[mode == "BUS", allow_same_route_transfer := FALSE]
```

Checking the results below, everything looks OK:

### Transfer fares configuration

@@

-   The fare for "BUS" to "BUS" integration is composed of 480 for the first leg plus 240 for the second leg, which equals to a total fare of 720.

```{r}
fare_settings$fare_per_transfer[leg1 == "BUS" & leg2 == "BUS", 
                                fare := 720]
```

-   Transfers between "RAIL" and "RAIL" are free and unlimited, which is already accounted for in the field `unlimited_transfers` of the `fare_per_mode` table. Thus, the equivalent row of the `fare_per_transfer` data.frame needs to be removed. If we leave the that row in `fare_per_transfer`, transfers between "RAIL" and "RAIL" will count to the global `max_discounted_transfers` allowance.

@@

Now, we can answer questions like "how many jobs one can access in 60 minutes using public transport, on a \$5 budget?".

@@

# helper function to calculate accessibility with common settings
calculate_accessibility <- function(settings, fare) {
  if (!is.null(settings)) {
    debug_file <- here::here("tests_marcus/debug_poa", 
                             paste0("debug_fare_", fare, ".csv" ))
    settings$debug_settings$output_file = debug_file
  }

  access_df <- accessibility(r5r_core,
                        origins = points_small,
                        destinations = points_small,
                        departure_datetime = departure_datetime,
                        opportunities_colname = "unit",
                        mode = c("WALK", "TRANSIT"),
                        cutoffs = c(60),
                        fare_calculator_settings = settings,
                        max_fare = fare,
                        max_trip_duration = 60,
                        max_walk_dist = 800,
                        time_window = 1,
                        percentiles = 50,
                        verbose = FALSE)
}
```

```{r}
access_unl <- calculate_accessibility(NULL, 0)
access_300 <- calculate_accessibility(fare_settings, 300)
access_500 <- calculate_accessibility(fare_settings, 500)
access_900 <- calculate_accessibility(fare_settings, 900)
```

```{r}
access_unl$max_fare <- "unl"
access_300$max_fare <- "3.0"
access_500$max_fare <- "5.0"
access_900$max_fare <- "9.0"

access_df <- rbind(access_unl, access_300, access_500, access_900)
```

```{r}
access_df$geometry <- h3jsr::h3_to_polygon(access_df$from_id)
access_df <- st_as_sf(access_df, crs = 4326)
```

```{r}
access_df %>%
  filter(cutoff == 60) %>%
  ggplot() +
  geom_sf(aes(fill=accessibility), color = NA) +
  scale_fill_distiller(palette = "Spectral") +
  facet_wrap(~max_fare)
```

```{r}
debug_df <- read_csv(here::here("tests_marcus/debug_poa", "debug_fare_300.csv"))
debug_df <- read_csv(here::here("tests_marcus/debug_poa", "debug_fare_500.csv"))
debug_df <- read_csv(here::here("tests_marcus/debug_poa", "debug_fare_900.csv"))
```

```{r}
access_df %>%
  st_set_geometry(NULL) %>%
  pivot_wider(names_from = max_fare, values_from = accessibility) %>%
  janitor::clean_names()
```

```{r}
write_fare_calculator(fare_settings, file_path = here::here("teste_fares_poa.zip"))
fare_settings_2 <- read_fare_calculator(file_path = here::here("teste_fares_poa.zip"))
```
